---
title: "ps5"
author: "Nicholas Mendoza"
date: "2023-02-20"
output: html_document
---
```{r}
library(tidyverse)
library(dplyr)
```

## Part 1: Load and check data
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.
created a newmarkdown file
2. (2pt) Load data. How many rows/columns do we have?
```{r}
gm <- read_delim("gapminder.csv.bz2")
dim(gm)
```
3. (2pt) Print a small sample of data. Does it look OK?
```{r}
gm %>% 
  sample_n(5)
```

## Part 2: Descriptive Statistics
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name
```{r}
gm %>% 
  group_by(name) %>%   
  filter(!is.na(name)) %>% 
  summarise(n = n_distinct(name)) %>% 
  arrange(desc(n))
```
There are 249 unique country names.
```{r}
gm %>% 
  group_by(iso2) %>%   
  filter(!is.na(iso2)) %>% 
  summarise(n=n_distinct(iso2)) %>% 
  arrange(desc(n))
```
There are 248 unique country iso2 codes.
```{r}
gm %>% 
  group_by(iso3) %>%   
  filter(!is.na(iso3)) %>% 
  summarise(n=n_distinct(iso3)) %>% 
  arrange(desc(n))
```
There are 253 unique country iso3 codes.

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?

```{r}
gm %>% 
  group_by(name) %>% 
  filter(is.na(iso2)) %>% 
  summarize(n = n_distinct(iso2)) %>% 
  arrange(desc(n))
```
Namibia does not have an iso2 code and there is another country that does not have a country name, just NA

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
gm %>% 
  group_by(iso3) %>% 
  filter(is.na(name)) %>% 
  summarize(n = n_distinct(name)) %>% 
  arrange(desc(n))
```
CHANISL, GBM, KOS, and NLD_CURACAO have iso3 codes, but not a name

3. (2pt) What is the minimum and maximum year in these data?
```{r}
min(gm$time, na.rm = TRUE)
max(gm$time, na.rm = TRUE)
```
1960 is the minimum year; 2019 is the maximum year.

## Part 3: CO2 emissions
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
gm %>% 
  filter(iso3 ==c("USA","CHN","IND","ZWE", "ARG")) %>% 
  ggplot() +
  geom_line(aes(x=time,y=co2,col=iso3)) +
  geom_point(aes(x=time,y=co2,col=iso3))
```
I notice that countries that are more developed have more CO2 emissions. You can also see that recently developed nations,
like India and China, have had an increase in CO2 emissions over time as they developed.

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
gm %>% 
  filter(iso3 ==c("USA","CHN","IND","ZWE", "ARG")) %>% 
  ggplot() +
  geom_line(aes(x=time,y=co2_PC,col=iso3)) +
  geom_point(aes(x=time,y=co2_PC,col=iso3))
```
This figure suggests that the USA is by far the leader in CO2 emissions per capita. All other nations are
towards the bottom of the graph. This may be because that US is significantly more developed than the other
countries listed, as it was one of the key leaders in the industrial revolution.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.

Here are the average CO2 emissions per capita in the year 2016.
```{r}
gm %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC), !is.na(region), time == 2016) %>% 
  summarise(mean = mean(co2_PC)) %>% 
  arrange(desc(mean))
```

Here are the average CO2 emissions per capita for all time.
```{r}
gm %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>% 
  summarise(mean = mean(co2_PC)) %>% 
  arrange(desc(mean))
```

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

```{r}
gm %>% 
  filter(time %in% c(1960, 2016), !is.na(region), !is.na(co2_PC)) %>% 
  group_by(time, region) %>% 
  summarise(avg_co2PC = mean(co2_PC)) %>% 
  ggplot(aes(x=region,y=avg_co2PC, fill=as.factor(time))) +
  geom_col(position = "dodge") +
    labs(title = "Average CO2 emissions per capita by year and region",
         x="region",
         y="average co2 emissions per capita") +
  scale_fill_discrete(name= "Year")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

Because the 2019 data is missing, I will take the 2016 data to answer this question

3 smallest
```{r}
gm %>% 
  filter(time == "2016", !is.na(region), !is.na(co2_PC)) %>% 
  group_by(region) %>% 
  arrange(co2_PC) %>% 
  slice_head(n=3) %>% 
  select(region, name, co2_PC)
```

3 largest
```{r}
gm %>% 
  filter(time == "2016", !is.na(region), !is.na(co2_PC)) %>% 
  group_by(region) %>% 
  arrange(co2_PC) %>% 
  slice_tail(n=3) %>% 
  select(region, name, co2_PC) 
```

## Part 4: GDP per capita
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960.
Make the point size dependent on the country size, and color those according to the continent.
Feel free to adjust the plot in other ways to make it better.

```{r}
gm %>% 
  filter(time == 1960, !is.na(region)) %>% 
  ggplot(aes(x=GDP_PC,y=lifeExpectancy,col=region,size=totalPopulation)) +
  geom_point()
  
```

2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
gm %>% 
  filter(time == 2019, !is.na(region)) %>% 
  ggplot(aes(x=GDP_PC,y=lifeExpectancy,col=region,size=totalPopulation)) +
  geom_point()
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

Both GDP per capita and life expectancy have increased for all countries through the past 60 years.
These factors seem to be related to some degree, as a higher GDP per capita represents a greater
life expectancy. However, there are diminishing returns with GDP per capita and life expectancy.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?

```{r}
gm %>% 
  filter(!is.na(region), !is.na(lifeExpectancy), !is.na(time)) %>% 
  filter(time %in% c("1960","2019")) %>% 
  group_by(region, time) %>% 
  summarize(average = mean(lifeExpectancy))
```
Yes, the results generally align with the countries from the regions in the figures in terms of life expectancy.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
```{r}
gm %>% 
  filter(!is.na(region), !is.na(lifeExpectancy), !is.na(time)) %>% 
  filter(time %in% c("1960","2019")) %>% 
  group_by(region, time) %>% 
  summarize(average = mean(lifeExpectancy)) %>% 
  mutate(prev = lag(average), growth = average - prev) %>% 
  filter(!is.na(growth)) %>% 
  arrange(desc(growth))
```
Continents with a lot of countries that are still in the developing phase have seen a greater increase in life
expectancy while continents that have countries that have already been developed for a long time have a lower
increase in life expectancy. For example, many countries in Africa and Asia are not fully developed yet and therefore
contribute to Africa and Asia's great increase in life expectancy while countries in Europe have already been developed
for a long time, demonstrated by their low increase in life expectancy.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
gm %>% 
  filter(time %in% c(1960, 2019), !is.na(GDP_PC)) %>% 
  ggplot(aes(x=GDP_PC, fill = factor(time))) +
  geom_histogram(alpha = 0.5, position = "dodge", bins = 30) +
  scale_fill_manual(values = c("red", "blue"), labels = c("1960" , "2019")) +
  labs(x = "GDP per capita", y= "Count", title = "GDP per capita for 1960 and 2019")
```

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.

1960 life expectancy rank
```{r}
gm %>% 
  filter(time == "1960", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(rank)
```
2019 life expectancy rank
```{r}
gm %>% 
  filter(time == "2019", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(rank)
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.



